---
title: "Analysis_BESHAY_PNB3EE3"
author: "Justin Beshay"
date: "2025-03-26"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Load required packages
library(ggplot2)
library(readr)
library(stats)
library(tidyverse)
library(dplyr)
library(rstatix)
library(ggpubr)
library(effectsize)
```

## 1. Introduction

This analysis examines whether different background noise conditions (Silence, White Noise, Music) affect memory recall performance. The primary dependent variable is recall accuracy. Secondary variables include recall time, music familiarity (music condition only), and study habits.

## 2. Simulate Data

```{r}
set.seed(123) # Set seed for reproducibility
sample_size <- 177
conditions <- c("silence", "white_noise", "music")
music_habits <- c("Yes", "No")

# Randomly assign participants to conditions
condition_assignments <- rep(conditions, each = sample_size / 3)
condition_assignments <- sample(condition_assignments, replace = FALSE)

# Create initial data frame
# MusicFamiliarity is only meaningful for the music condition
# RecallAccuracy and RecallRT are generated with realistic means & SDs per condition
df <- data.frame(
  Participant = 1:sample_size,
  Condition = condition_assignments,
  RecallAccuracy = NA,
  RecallRT = NA,
  MusicFamiliarity = NA,
  MusicHabit = sample(music_habits, size = sample_size, replace = TRUE),
  stringsAsFactors = TRUE
)
# Generate scores by condition
for (i in 1:nrow(df)) {
  cond <- df$Condition[i]
  if (cond == "silence") {
    df$RecallAccuracy[i] <- round(rnorm(1, mean = 75, sd = 10), 1)
  } else if (cond == "white_noise") {
    df$RecallAccuracy[i] <- round(rnorm(1, mean = 65, sd = 10), 1)
  } else if (cond == "music") {
    df$RecallAccuracy[i] <- round(rnorm(1, mean = 55, sd = 10), 1)
    df$MusicFamiliarity[i] <- sample(1:10, 1)
  }
  df$RecallRT[i] <- round(rnorm(1, mean = 50, sd = 10), 1)
}
# Save for loading later
write.csv(df, "PNB3EE3_simulated_data_177.csv", row.names = FALSE)
```

## 3. Clean and Format Data

```{r}
df <- read.csv("PNB3EE3_simulated_data_177.csv") # Load and prepare data

df$Condition <- as.factor(df$Condition)
df$MusicHabit <- as.factor(df$MusicHabit)
df$MusicFamiliarity <- as.numeric(df$MusicFamiliarity)
df$RecallAccuracy <- pmin(pmax(df$RecallAccuracy, 0), 100) # Clip recall accuracy to [0,100] in case of outliers
```

## 4. Descriptive Statistics

```{r}
# Overall Descriptive Stats
summary(df)

# Recall Accuracy by Condition, # Mean, SD, Min, Max Recall Accuracy by Condition
df %>%
  group_by(Condition) %>%
  summarise(
    Mean_Recall = round(mean(RecallAccuracy, na.rm = TRUE), 2),
    SD_Recall = round(sd(RecallAccuracy, na.rm = TRUE), 2),
    Min_Recall = round(min(RecallAccuracy, na.rm = TRUE), 2),
    Max_Recall = round(max(RecallAccuracy, na.rm = TRUE), 2),
    n = n()
  )

# Recall Response Time by Condition
df %>%
  group_by(Condition) %>%
  summarise(
    Mean_RT = round(mean(RecallRT, na.rm = TRUE), 2),
    SD_RT = round(sd(RecallRT, na.rm = TRUE), 2)
  )

# Music Familiarity (Music Only)
df %>%
  filter(Condition == "music") %>%
  summarise(
    Mean_Familiarity = round(mean(MusicFamiliarity, na.rm = TRUE), 2),
    SD_Familiarity = round(sd(MusicFamiliarity, na.rm = TRUE), 2)
  )

# Study Habit Breakdown
df %>%
  group_by(MusicHabit) %>%
  summarise(
    MeanRecall = round(mean(RecallAccuracy), 2),
    SDRecall = round(sd(RecallAccuracy), 2),
    Count = n()
  )
```

## 5. Exploratory Visualizations

### 5.1 Scatter: Familiarity vs Recall (Music)

```{r}
# Scatterplot: Music Familiarity vs Recall Accuracy
## Secondary Hypothesis: Is more familiar music more distracting?

df %>%
  filter(Condition == "music") %>%
  ggplot(aes(x = MusicFamiliarity, y = RecallAccuracy)) +
  geom_point(color = "purple", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Figure 1. Music Familiarity vs Recall Accuracy (Music Condition)",
    x = "Music Familiarity (1–10)",
    y = "Recall Accuracy (%)"
  )
```

### 5.2 Scatter: Recall Time vs Accuracy

```{r}
#Scatterplot: Recall Time vs Recall Accuracy
##Exploratory: Are slower participants more accurate (or not)?

ggplot(df, aes(x = RecallRT, y = RecallAccuracy)) +
  geom_point(color = "darkred", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Figure 2. Recall Time vs Recall Accuracy (All Participants)",
    x = "Recall Time (seconds)",
    y = "Recall Accuracy (%)"
  )
```

### 5.3 Study Habit Distribution

```{r}
# Bar Plot: Music Study Habit Frequencies
##Demographics: How many people study with music?


ggplot(df, aes(x = MusicHabit, fill = MusicHabit)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Figure 3. Frequency of Participants Who Study with Music",
    x = "Do You Listen to Music While Studying?",
    y = "Number of Participants"
  )
```

### 5.4 Boxplot: Recall by Study Habit (Music Group)

```{r}
# Boxplot: Recall Accuracy by Music Habit (Music Condition Only)
## Exploratory Moderation: Do music study habits protect recall?

df %>%
  filter(Condition == "music") %>%
  ggplot(aes(x = MusicHabit, y = RecallAccuracy, fill = MusicHabit)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Figure 4. Recall Accuracy by Music Study Habit (Music Condition Only)",
    x = "Studies with Music?",
    y = "Recall Accuracy (%)"
  )
```

## 6. Inferential Analyses

### 6.1 ANOVA

```{r}
# Test for overall group differences in recall accuracy across the three conditions
anova_model <- aov(RecallAccuracy ~ Condition, data = df)
summary(anova_model)
eta_squared(anova_model) # Calculate effect size (eta squared)
```

### 6.2 One-Tailed T-Tests

```{r}
# Planned comparisons to follow up the ANOVA
# Test whether silence > music
t.test(RecallAccuracy ~ Condition, data = df %>% filter(Condition %in% c("silence", "music")), alternative = "less")

# Test whether silence > white noise
t.test(RecallAccuracy ~ Condition, data = df %>% filter(Condition %in% c("silence", "white_noise")), alternative = "greater")

# Test whether white noise > music
t.test(RecallAccuracy ~ Condition, data = df %>% filter(Condition %in% c("white_noise", "music")), alternative = "less")
```

### 6.3 Cohen’s d

```{r}
# Estimate effect size for the difference between silence and music
cohens_d(RecallAccuracy ~ Condition, data = df %>% filter(Condition %in% c("silence", "music")))
```

### 6.4 Correlation: Familiarity × Recall

```{r}
# Test whether recall accuracy is associated with music familiarity within the music group
cor.test(df$RecallAccuracy[df$Condition == "music"], df$MusicFamiliarity[df$Condition == "music"])
```

### 6.5 Condition × Study Habit Interaction

```{r}
# Linear model testing interaction between condition and music study habit
interaction_model <- lm(RecallAccuracy ~ Condition * MusicHabit, data = df)
summary(interaction_model)
```

## 7. Final Visuals (Planned Visualizations)

```{r}
# Boxplot comparing recall accuracy distributions across the three conditions
ggplot(df, aes(x = Condition, y = RecallAccuracy, fill = Condition)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Figuire 5. Recall Accuracy by Background Noise Condition")

# Bar plot of means with standard error bars
summary_stats <- df %>%
  group_by(Condition) %>%
  summarise(
    mean_recall = mean(RecallAccuracy),
    se_recall = sd(RecallAccuracy) / sqrt(n())
  )

ggplot(summary_stats, aes(x = Condition, y = mean_recall, fill = Condition)) +
  geom_col(width = 0.6, colour = "black") +
  geom_errorbar(aes(ymin = mean_recall - se_recall, ymax = mean_recall + se_recall), width = 0.2) +
  theme_minimal() +
  labs(title = "Figuire 6. Mean Recall Accuracy by Condition ± Standard Error", y = "Recall Accuracy (%)")
```

```{r}
# Histogram: Recall Accuracy (All Participants)
## Assumption Check: Is recall accuracy approximately normal?
# Histogram of recall accuracy to check distribution shape (normality assumption)
ggplot(df, aes(x = RecallAccuracy)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Figure 7. Distribution of Recall Accuracy (All Conditions)",
    x = "Recall Accuracy (%)",
    y = "Count"
  )
```
```{r}
#Histogram: Recall Response Time
## Assumption Check & Exploratory Variable (Response Speed)
# Histogram of response time to check its distribution
ggplot(df, aes(x = RecallRT)) +
  geom_histogram(binwidth = 5, fill = "darkgreen", color = "black") +
  theme_minimal() +
  labs(
    title = "Figure 8. Distribution of Recall Response Time",
    x = "Recall Time (seconds)",
    y = "Count"
  )
```
```{r}
#Histogram: Music Familiarity (Music Condition Only)
##Exploratory: Distribution of familiarity ratings (1–10)

df %>%
  filter(Condition == "music") %>%
  ggplot(aes(x = MusicFamiliarity)) +
  geom_histogram(binwidth = 1, fill = "orchid", color = "black") +
  theme_minimal() +
  labs(
    title = "Figure 9. Distribution of Music Familiarity Ratings (Music Condition)",
    x = "Familiarity Rating (1–10)",
    y = "Count"
  )
```


## 8. Conclusion
- The silence condition produced the highest recall accuracy, supporting the primary hypothesis that background noise negatively affects memory performance. As shown in Figure 6, participants in the silence condition had a mean recall accuracy of 74.47%, which was significantly higher than both the white noise group (M = 65.91%) and the music group (M = 56.33%). Although the standard error bars in Figure 5 show some overlap, When including standard error, the bars dont overlap (Figuire 6). These differences were statistically confirmed by a one-way ANOVA (F(2, 174) = 52.26, p < .001, η² = .38) and follow-up t-tests:

  - Silence > Music: t(114.19) = –10.35, p < .001 (one-tailed)

  - Silence > White Noise: t(114.75) = 4.94, p < .001 (one-tailed)

  - White Noise > Music: t(115.95) = –5.21, p < .001 (one-tailed)

- Contrary to expectations, music familiarity was not significantly correlated with recall accuracy within the music condition (r = .11, p = .42), suggesting that greater familiarity with a song did not meaningfully impact memory performance.

- An exploratory interaction model was used to test whether music study habits moderated the effect of background noise on recall accuracy. The overall model was statistically significant, F(5, 171) = 21.68, *p* < .001, with an R² of .388, meaning the model explained about 39% of the variance in recall performance. However, the interaction terms were not statistically significant:

  - Silence × MusicHabit: β = –0.53, p = .88  
  - White Noise × MusicHabit: β = 3.49, p = .34  

  - These p-values are well above .05, suggesting that the impact of background noise on recall was consistent regardless of whether participants usually study with music.

- Future research should expand on these findings by including delayed recall, directly manipulating music familiarity or complexity, and comparing lyrical vs. instrumental music to better understand the specific features of music that disrupt or facilitate memory.



