---
title: "My Pre-Registration"
output:

  md_document: 
        variant: markdown_github
  html_document: default
  pdf_document: default
---

# Pre-registration Template
Adapted from [https://osf.io/preprints/metaarxiv/epgjd](https://osf.io/preprints/metaarxiv/epgjd) into markdown format.

<details>

-   [Study Information](#study-information)
    -   [1. Title](#1-title)
    -   [2. Authors](#2-authors)
    -   [3. Description](#3-description)
    -   [4. Hypotheses](#4-hypotheses)
-   [Design Plan](#design-plan)
    -   [5. Study type](#5-study-type)
    -   [6. Blinding](#6-blinding)
    -   [7. Is there any additional blinding in this study?](#7-is-there-any-additional-blinding-in-this-study-)
    -   [8. Study design](#8-study-design)
    -   [9. Randomization](#9-randomization)
-   [Sampling Plan](#sampling-plan)
    -   [10. Existing data](#10-existing-data)
    -   [11. Explanation of existing data](#11-explanation-of-existing-data)
    -   [12. Data collection procedures](#12-data-collection-procedures)
    -   [13. Sample size](#13-sample-size)
    -   [14. Sample size rationale](#14-sample-size-rationale)
    -   [15. Stopping rule](#15-stopping-rule)
-   [Variables](#variables)
    -   [16. Manipulated variables](#16-manipulated-variables)
    -   [17. Measured variables](#17-measured-variables)
    -   [18. Indices](#18-indices)
-   [Analysis Plan](#analysis-plan)
    -   [19. Statistical models](#19-statistical-models)
    -   [20. Transformations](#20-transformations)
    -   [21. Inference criteria](#21-inference-criteria)
    -   [22. Data exclusion](#22-data-exclusion)
    -   [23. Missing data](#23-missing-data)
    -   [24. Exploratory analysis](#24-exploratory-analysis)
-   [Other](#other)
    -   [25. Other](#25-other)

</details>

# Study Information 

## 1. Title 

Silence vs. Sound: Investigating How Background Music and Noise Influence Study Retention

## 2. Authors 

Justin Beshay

## 3. Description 


Memory retention is a fundamental aspect of learning, and environmental factors such as background noise may influence recall accuracy. However, prior research has produced mixed findings on whether background noise—particularly music—enhances or impairs memory performance. Some studies suggest that background music can improve cognitive performance by enhancing mood, increasing arousal, and reducing stress, which may facilitate attention and encoding processes [@carr2016; @mammarella2007; @taheri2022]. Others argue that background music can be distracting, as it competes for attentional resources needed for memory encoding and retrieval, leading to impaired recall accuracy [@lehmann2017; @echaide2019]. Given these conflicting findings, further research is needed to clarify how different types of background noise influence memory retention.

One major limitation in the existing literature is the variability in study designs, including differences in music type, volume, complexity, and familiarity. For example, studies using instrumental music have sometimes found no effect or even benefits for memory, whereas music with lyrics has been shown to disrupt verbal processing tasks [@souza2023]. The familiarity of the music may also be a key factor—highly familiar songs may induce involuntary engagement, such as singing along or anticipating lyrics, which could increase cognitive load and reduce memory performance [@chew2016]. Additionally, the complexity of the music (e.g., fast tempo, unpredictable melodies) may influence its impact on recall, with high-complexity music likely being more disruptive than simple, ambient sounds [@gonzalez2019]. These inconsistencies highlight the need for a systematic investigation that directly compares the effects of different background noise conditions under controlled experimental conditions.

Another important gap in the literature is the role of individual differences in study habits. Some students habitually study with music, while others prefer silence, raising the question of whether Frequent music listeners are less distracted by background music while studying compared to those who usually study in silence [@furnham2002]. If frequent music listeners develop adaptations to studying with noise, their recall performance may be less negatively impacted than those who are accustomed to studying in quiet environments. However, this remains an open question, as relatively few studies have considered study habits as a moderating variable in memory research when interacting with music/noise [@christopher2017].  


The present study aims to clarify these mixed findings by systematically examining how three background noise conditions—silence, white noise, and music—impact recall accuracy. Specifically, we will test whether music impairs recall more than silence and white noise, and whether the effect of music depends on factors such as song familiarity and complexity. We will also explore whether individual study habits moderate the effects of background noise on memory, with habitual music listeners potentially showing greater resistance to distraction.

To address these research gaps, participants will be randomly assigned to one of three conditions (Silence, White Noise, or Music) and will study a list of words. After a 10-minute delay, they will complete a free recall test. Participants in the Music Condition will also rate their familiarity with the song to assess its potential moderating effect on recall accuracy.

Findings from this study may have important implications for optimizing study environments, particularly for students who listen to music while studying. If background noise impairs memory performance, educators may need to reconsider best practices for study settings. Alternatively, if specific types of music enhance memory retention, targeted use of background noise could serve as a cognitive aid in educational contexts.



## 4. Hypotheses 
1. Primary Hypothesis (Directional): Participants who study in the silence condition will have higher recall accuracy compared to those in the music and white noise conditions.

2. Secondary Hypothesis (Directional): Participants in the music condition will have lower recall accuracy than those in the silence and white noise condition, but the effect will depend on the complexity and familiarity of the music.
- **High-complexity music** (e.g., fast-paced instrumental music, songs with unpredictable melodies, or multiple layers of sound) will impair recall accuracy more than low-complexity music (e.g., slow instrumental or ambient tracks).
- **Familiar music** will be more distracting than unfamiliar music, as participants may involuntarily sing along or anticipate lyrics, leading to greater cognitive interference with memory encoding and retrieval.
- **Music with lyrics** will be more disruptive to verbal memory tasks than instrumental music, as processing lyrics competes with the cognitive resources required for studying and recall.

  
3. Long-Term Retention Hypothesis (Directional): Participants who study in the silence condition will exhibit better long-term retention of studied material compared to those in the music and white noise conditions.

4. Interaction Hypothesis: The impact of background noise on recall accuracy will vary depending on individual differences in study habits (e.g., whether participants frequently listen to music while studying).
- Participants who frequently listen to the same songs while studying may experience less distraction from familiar music compared to those who typically study in silence.



# Design Plan 


## 5. Study type 

This study qualifies as an experiment because it involves direct manipulation of an independent variable—background noise condition—and measures its effect on a dependent variable, recall accuracy. Participants will be randomly assigned to one of three conditions (silence, white noise, or music), ensuring that differences in memory performance are due to the experimental manipulation rather than pre-existing individual differences. By controlling for extraneous variables, such as study time and recall delay, and using a standardized procedure across conditions, this study allows for causal inferences about how different types of background noise influence memory retention. The structured design, with controlled exposure times and a fixed delay period before recall, makes it a laboratory-style experiment conducted in a digital setting using jsPsych.

## 6. Blinding 

- [ ] No blinding is involved in this study.

- [ ] For studies that involve human subjects, they will not know the treatment group to which they have been assigned.

- [ ] Personnel who interact directly with the study subjects (either human or non-human subjects) will not be aware of the assigned treatments. (Commonly known as “double blind”)

- [x] Personnel who analyze the data collected from the study are not aware of the treatment applied to any given group.

## 7. Is there any additional blinding in this study? 

This study is a single-blind experiment, meaning that participants will be aware of their assigned condition, but the researchers analyzing the data will not know which condition each participant was in.

Participants must be aware of their assigned condition because the independent variable—background noise—directly affects their sensory experience during the study phase. Unlike in pharmaceutical or placebo studies where conditions can be masked, participants actively hear either silence, white noise, or music while studying. Since it is impossible to expose participants to different noise conditions without them noticing, they are inherently aware of their assigned condition. However, to prevent potential bias in data analysis, group assignments (white noise, music, or silence) will concealed from researchers until all statistical analyses are completed.


## 8. Study design 

We have a between-subjects design with one factor (background noise condition) with three levels: Silence, White Noise, and Music. Participants will be randomly assigned to one of these conditions, and recall accuracy will be measured as the percentage of correctly recalled words. Since each participant experiences only one condition, no counterbalancing is required.


## 9. Randomization 

This study will use block randomization to assign participants to one of three background noise conditions (silence, white noise, or music). Block randomization ensures that the number of participants in each condition remains balanced throughout the study. Each block will contain an equal number of participants assigned to one of the three conditions. Random assignment will be implemented automatically using a randomization function within jsPsych.

# Sampling Plan 

## 10. Existing data 

- [x] Registration prior to creation of data: As of the date of submission of this research plan for preregistration, the data have not yet been collected, created, or realized.

## 11. Explanation of existing data 

This study does not use any pre-existing data. All data will be collected after preregistration, ensuring that confirmatory analyses are specified before any observations are made. All analyses will be conducted on newly collected data. 

## 12. Data collection procedures 

### Population & Recruitment
Participants will be undergraduate students recruited through SONA McMaster University’s research participation system. Recruitment will be conducted via SONA study postings, and participants will receive course credit for their participation.

### Eligibility Criteria
- Age: 18+  
- Language: Fluent in English  
- Hearing: Normal or corrected-to-normal hearing  
- Device: Must use a computer 
- Completion: Must finish the study in one session; incomplete responses will be excluded  

### Study Timeline
1. Informed Consent & Demographics:
   - Participants provide consent and complete a demographic survey.  

2. Study Phase:  
   - Participants are randomly assigned to one of three background noise conditions:  
     - Silence  
     - White Noise  
     - Music  
   - They study a word list.  

3. Delay & Distractor Task: 
   - A 10-minute delay with a neutral distractor task follows.  

4. Recall Test:
   - Participants recall as many words as possible.  

5. Music Familiarity Rating: (Music condition only)
   - Participants rate song familiarity on a 1-10 scale.  

6. Debriefing & Credit Allocation:  
   - Participants receive a debriefing statement and SONA credit.  

### Data Collection Duration
Data collection will continue until the target sample size (determined via power analysis) is met.


## 13. Sample size 

the target sample size for this study is around 185 participants, with 59 participants per condition (silence, white noise, and music). Since this is a between-subjects design, each participant will be exposed to only one condition.

This sample size was determined using a power analysis for a one-way ANOVA, assuming an effect size of 0.3, a significance level of 0.05, and a desired power of 0.95. This ensures that the study has sufficient statistical power to detect meaningful differences between the conditions.

To account for potential dropouts or incomplete responses we rounded up from 177  participants (directed from the power analysis) to 185, slightly more participants than the minimum required. Only participants who fully complete the study will be included in the final analysis.

## 14. Sample size rationale 

```{r}
# Load necessary package
options(repos = "https://cran.rstudio.com")  # Set default CRAN mirror
install.packages("pwr")  
install.packages("pwr")  
library(pwr)

# Set parameters
effect_size <- 0.3  # Cohen's f
alpha <- 0.05       # Significance level
power <- 0.95       # Desired power
num_groups <- 3     # Number of groups

# Perform power analysis
sample_size_per_group <- pwr.anova.test(k = num_groups, 
                                        f = effect_size, 
                                        sig.level = alpha, 
                                        power = power)$n
(sample_size_per_group) # = 58.2 so round to 59 participants per group

# Calculate total sample size
total_sample_size <- sample_size_per_group * num_groups

# Print results
print("Sample size per group: 59")
print("Total sample size: 59 x 3 = 177")
```

## 15. Stopping rule 

Sign-ups will remain open on SONA until we reach 185 completed participants, as determined in #13 and #14 Sample Size.


# Variables 


## 16. Manipulated variables 

The manipulated variable (IV) in this study is background noise condition, which has three levels:

1. Silence Condition – No background noise will be played during the study phase. Participants will study in a quiet environment, with only system-generated instructions presented.
2. White Noise Condition – A continuous, non-rhythmic white noise will be played at a standardized volume (e.g., 50 dB) throughout the study phase. This will be a pre-recorded white noise track that remains consistent for all participants.
3. Music Condition – A song with lyrics will be played at a standardized volume (e.g., 50 dB) during the study phase. The song will be the same for all participants in this condition to maintain consistency.") 

The background noise will be played through their device speakers or headphones. Participants will not have control over playback



## 17. Measured variables 

The primary dependent variable in this study is recall accuracy, measured as the percentage of correctly recalled words from the studied word list. Participants will be asked to type all the words they remember after a 10-minute delay following a distraction task.

Additional measured variables include:

1. Response Time (Exploratory Measure): The time taken by participants to recall and enter words will be recorded to explore potential differences in retrieval speed across conditions.
2. Study Habits (Covariate): Participants will self-report whether they typically listen to music while studying, which may moderate the effects of background noise on recall accuracy.
3. Music Familiarity (Exploratory Measure): If assigned to the music condition, participants will indicate whether they were familiar with the song played.


## 18. Indices 
In this study, the primary measure of interest is recall accuracy, which will be computed as the percentage of correctly recalled words from the studied list.
```{r}
#correct_recalled <- sum(#participant_response %in% study_list)

#total_words <- 20
#recall_accuracy <- (correct_recalled / total_words) * 100

```
Once we collect data from jsPsych, we can determine correctly recalled words by comparing the participant’s response with the original word list.

2. Mean Recall Time
Measures how long participants take to recall words before submitting their response.
This is used to find out if recall time is related to recall accuracy 
```{r}
#total_recall_time <- (final_submission_time - recall_start_time)
```


3. Music Familiarity (Self-reported familiarity rating (1-10))
Used to explore whether familiar music is more distracting than unfamiliar music. Participants in the Music Condition will rate their familiarity with the song on a Likert scale (1 to 10): 1 = Not familiar at all, 10 = Extremely familiar

```{r}
#music_familiarity_index <- ifelse(condition == "Music", as.numeric(familiarity_rating), NA)

## (Ensures only participants in the Music Condition have a familiarity score; others get NA.)
```


# Analysis Plan 


## 19. Statistical models 

We will use a **one-way between-subjects ANOVA** to test the effect of **background noise** on **recall accuracy**.

- **Independent Variable (IV):** Background noise condition (**Silence, White Noise, Music**)  
- **Dependent Variable (DV):** Recall accuracy (**percentage of correctly recalled words**)  

### **Primary Analysis**
- A **one-way ANOVA** will determine whether background noise condition significantly affects **recall accuracy**, testing the **Primary Hypothesis** that **participants in the silence condition will recall more words** than those in the **music and white noise conditions**.
- **Planned one-tailed t-tests** will compare recall accuracy between conditions based on our **directional hypotheses**:  
  - **Silence > Music** (higher recall accuracy in silence than in music).  
  - **Silence > White Noise** (higher recall accuracy in silence than in white noise).  
  - **White Noise > Music** (music impairs recall more than white noise).  

### **Secondary Analysis**
To examine the **Secondary Hypothesis**, we will explore whether:  
- **Music complexity** affects recall accuracy (e.g., high-complexity music, such as unpredictable melodies, will be more disruptive).  
- **Music familiarity** moderates recall performance (e.g., familiar music may be more distracting due to involuntary engagement).  
- **Music with lyrics** impairs recall more than instrumental music, due to interference with verbal encoding.  

These analyses will use **interaction effects** and **correlation tests** to assess the impact of **familiarity, complexity, and lyrics presence** in the **music condition** only.  

### **Long-Term Retention Hypothesis**  
If a **delayed recall phase** is included, we will use **repeated-measures ANOVA** to test whether participants who studied in **silence** exhibit **better long-term retention** than those in the **music and white noise conditions**.  

### **Interaction Hypothesis & Exploratory Analyses**  
To assess whether **study habits** influence recall performance (**Interaction Hypothesis**), we will examine:  
- Whether **habitual music listeners** perform differently in the **music condition** compared to participants who typically study in silence.  
- Whether **participants who frequently listen to the same songs while studying** experience less distraction from familiar music.  

These relationships will be tested using **moderation analyses** (e.g., ANCOVA with study habits as a covariate) and **correlation tests** to explore how background noise effects vary based on **individual differences**.  

### **Effect Size & Reporting**
- **Cohen’s \( \eta^2 \)** will be reported for ANOVA to quantify effect size.  
- **Cohen’s \( d \)** will be used for planned pairwise t-tests.  
- **Correlation coefficients** (\( r \)) will be reported for relationships between music familiarity, complexity, and recall accuracy.  

All **exploratory findings** will be reported separately and labeled as such, ensuring transparency in distinguishing between **confirmatory and hypothesis-generating** analyses.



## 20. Transformations 

 The data from the effect of noise on recall accuracy dont require transformations. All variables are in their natural form and suitable for analysis. Categorical variables will be used as-is, and continuous variables such as recall accuracy and response time will be analyzed without normalization, centering, or scaling.


## 21. Inference criteria 

Primary Analysis: A one-way ANOVA will be conducted to test for differences in recall accuracy across the three conditions (silence, white noise, and music).


Statistical Significance: A p-value threshold of 0.05 (𝛼=0.05) will be used to determine statistical significance.


Post-hoc Comparisons: Since the hypotheses are directional, planned one-tailed t-tests will be conducted for pairwise comparisons instead of Tukey’s HSD.


Effect Size Reporting: Cohen’s 𝜂^2 (eta squared) will be reported for ANOVA, and Cohen’s 𝑑 will be reported for pairwise comparisons.


One-Tailed Tests: Given the directional nature of the hypotheses, all t-tests will be conducted as one-tailed tests, predicting that recall accuracy will be higher in silence than in music/white noise and that music will have a greater negative effect than white noise.


## 22. Data exclusion 

Non-Completion of Key Experimental Phases:
- Participants who fail to complete the study phase (e.g., do not view the full word list) will be removed.
- Participants who do not attempt to recall any words will be excluded, as their data does not contribute to memory performance analysis.

Invalid or Non-Compliant Responses:
- If participants enter nonsense words, repetitions, or random keyboard inputs during recall, their data will be flagged and reviewed for potential exclusion.

Outlier Handling:
- Participants with total recall times exceeding three standard deviations from the mean may be examined to determine if their data is valid.


## 23. Missing data 

Participants who do not complete the recall task (fail to submit any recalled words) will be excluded from the analysis, as their data would not contribute to assessing recall accuracy. Additionally, if a participant does not complete the study phase (e.g., exits before studying the full word list), their data will also be excluded.

For partial missing data:
If a participant recalls some words but does not submit within the allotted time, their recall accuracy will be calculated based on the words they provided before timeout.

## 24. Exploratory analysis 

Potential exploratory analyses include:

1. Effect of Study Habits on Recall Accuracy:
Participants will self-report whether they typically listen to music while studying. We will explore whether habitual music listeners perform differently in the music condition compared to those who usually study in silence.

2. Impact of Music Familiarity on Recall Performance (Music Condition Only):
Participants in the music condition will rate their familiarity with the song. We will explore whether familiarity affects recall accuracy (e.g., whether familiar music is more distracting than unfamiliar music).

3. Individual Differences in Response Time:
We will examine whether recall speed (total recall time) is associated with recall accuracy across conditions, potentially identifying patterns of fast vs. slow recallers.



# References

Carr, S. M., & Rickard, N. S. (2016). The use of emotionally arousing music to enhance memory for subsequently presented images. Psychology of Music, 44(5), 1145-1157. https://doi.org/10.1177/0305735615613846

Chew, A. S.-Q., Yu, Y.-T., Chua, S.-W., & Gan, S. K.-E. (2016). The effects of familiarity and language of background music on working memory and language tasks in Singapore. Psychology of Music, 44(6), 1431-1438. https://doi.org/10.1177/0305735616636209

Christopher, E. A., & Shelton, J. T. (2017). Individual differences in working memory predict the effect of music on student performance. Journal of Applied Research in Memory and Cognition, 6(2), 167–173. https://doi.org/10.1016/j.jarmac.2017.01.012

Echaide, C., del Río, D., & Pacios, J. (2019). The differential effect of background music on memory for verbal and visuospatial information. The Journal of General Psychology, 146(4), 443–458. https://doi.org/10.1080/00221309.2019.1602023

Furnham, A., & Strbac, L. (2002). Music is as distracting as noise: The differential distraction of background music and noise on the cognitive test performance of introverts and extraverts. Ergonomics, 45(3), 203–217. https://doi.org/10.1080/00140130210121932

Gonzalez, M. F., & Aiello, J. R. (2019). More than meets the ear: Investigating how music affects cognitive task performance. Journal of Experimental Psychology: Applied, 25(3), 431-444. https://doi.org/10.1037/xap0000202

Lehmann, J. A. M., & Seufert, T. (2017). The influence of background music on learning in the light of different theoretical perspectives and the role of working memory capacity. Frontiers in Psychology, 8, 1902. https://doi.org/10.3389/fpsyg.2017.01902

Mammarella, N., Fairfield, B., & Cornoldi, C. (2007). Does music enhance cognitive performance in healthy older adults? The Vivaldi effect. Aging Clinical and Experimental Research, 19(5), 394–399. https://doi.org/10.1007/BF03324720

Souza, A. S., & Leal Barbosa, L. C. (2023). Should we turn off the music? Music with lyrics interferes with cognitive tasks. Journal of Cognition, 6(1), 24. https://doi.org/10.5334/joc.273

Taheri, S., Razeghi, M., Choobineh, A., Kazemi, R., Rasipisheh, P., & Vali, M. (2022). Investigating the effect of background music on cognitive and skill performance: A cross-sectional study. Work, 71(4), 871–879. https://doi.org/10.3233/WOR-213631







